---
title: "California Wildfire Analysis"
author: "Benjamin Garcia, Alejandro Cepeda, Brandon Ly, Brandon Dora"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

------------------------------------------------------------------------

#### Data Cleaning:

All rows in active column are False because none of them are active
anymore. This information is not particularly useful so Benjamin removed
this column. The canonical url column provides the ends of a url, but no
mention of the website being references. There is not any real value so
Benjamin removed the canonical url column. County id is not particularly
helpful since longitude and latitude values are already present.
Benjamin removed the County id column. The column featured is mostly
false and the column final is mostly true. Fuel type is mostly null.
Benjamin removed featured, final and fuel type columns together. The
percent contained column is all 100's since all fires reported have been
contained. Benjamin removed the percent contained column. The public
data column has all TRUE values, so Benjamin removed it. The status
column all finalized value so Benjamin removed it. All of the values in
StructuresEvacuated were NA so Benjamin removed it. The UniqueId column
makes no sense so Benjamin removed it. The update column provides no
useful dates like started and extinguished so Benjamin removed it.

```{r}
library(readr)
fire_dataset_Cleaning <- read_csv("California_Fire_Incidents.csv")
#View(fire_dataset_Cleaning)

fire_dataset_Cleaning <- fire_dataset_Cleaning[,-2]
fire_dataset_Cleaning <- fire_dataset_Cleaning[,-6]
fire_dataset_Cleaning <- fire_dataset_Cleaning[,-9]
fire_dataset_Cleaning <- fire_dataset_Cleaning[,-14:-16]
fire_dataset_Cleaning <- fire_dataset_Cleaning[,-21]
fire_dataset_Cleaning <- fire_dataset_Cleaning[,-22]
fire_dataset_Cleaning <- fire_dataset_Cleaning[,-25]
fire_dataset_Cleaning <- fire_dataset_Cleaning[,-27]
fire_dataset_Cleaning <- fire_dataset_Cleaning[,-28:-29]
```

Benjamin moved the started column to be in front of the established
column for clarity. Benjamin moved the name column to be the first
column of the dataset for clarity. Benjamin moved the Longitude after
the Latitude for clarity.

```{r}
library(dplyr)
fire_dataset_Cleaning <- fire_dataset_Cleaning %>% relocate(Started, .before = Extinguished)
fire_dataset_Cleaning <- fire_dataset_Cleaning %>% relocate(Name, .before = AcresBurned)
fire_dataset_Cleaning <- fire_dataset_Cleaning %>% relocate(Longitude, .before = Location)
```

In many of the columns, there were both NA values and numeric values.
Benjamin decided to replace NA values with 0 since it is unclear what
the true value of these observations might be. Please consider
airtankers, crewsinvolved, dozers, engines, fatalities, helicopters,
injuries, personnel involved, structures damaged, structures destroyed,
structures threatened and water tenders with care.

```{r}
library(tidyr)
summary(fire_dataset_Cleaning$Started)
min(fire_dataset_Cleaning$Started)
min(fire_dataset_Cleaning$Extinguished)
summary(fire_dataset_Cleaning$Extinguished)
```

------------------------------------------------------------------------

#### California Wildfire Shiny Web App:

```{r echo=FALSE, message=FALSE}
cleanFireData <- function() {
  # Benjamin's cleaning script as a function for server use
  testFrame <- read.csv("California_Fire_Incidents.csv")
  testFrame <- testFrame[,-2]
  testFrame <- testFrame[,-6]
  testFrame <- testFrame[,-9]
  testFrame <- testFrame[,-14:-16]
  testFrame <- testFrame[,-21]
  testFrame <- testFrame[,-22]
  testFrame <- testFrame[,-25]
  summary(testFrame$StructuresEvacuated)
  testFrame <- testFrame[,-27]
  summary(testFrame$StructuresThreatened)
  testFrame <- testFrame[,-28:-29]
  testFrame <- testFrame %>% relocate(Started, .before = Extinguished)
  testFrame <- testFrame %>% relocate(Name, .before = AcresBurned)
  testFrame <- testFrame %>% relocate(Longitude, .before = Location)
  testFrame <- replace_na(testFrame, list(AirTankers = 0, CrewsInvolved = 0, Dozers = 0, Engines = 0, Fatalities = 0, Helicopters = 0, Injuries = 0, PersonnelInvolved = 0, StructuresDamaged = 0, StructuresDestroyed = 0, StructuresThreatened = 0, WaterTenders = 0))
  
  return(testFrame)
}
```

The following section is intended to interactively visualize the various
variables available. The variables that Alejandro wants to focus on are
the `AcresBurned` throughout the four seasons of the year. With the help
of module 11's material, the process of developing and deploying the
shiny web app was possible with the use of the `shiny` and `rsconnect`
packages.

```{r message=FALSE}
# libraries
#install.packages(c('shiny','shinythemes','thematic','tidyverse','lubridate','rsconnect'))
library(shiny)
library(shinythemes)
library(thematic)
library(tidyverse)
library(lubridate)
library(rsconnect)
```

Below, Alejandro has defined a UI with two separate tabs within the
panel:

-   The **scatter plot** will display the result of choosing between
    three variables, the number of **acres burned**, **fatalities**, and
    **injuries**. These variables will be options for the x-axis,
    y-axis, and dot size of the plot.

-   The **bar graph** will instead visualize any trends in the number of
    acres burned, along with other observations, per each year. These
    will be separated by the season in which they occurred.

```{r}
# Define UI
ui <- fluidPage(
  theme = shinytheme("yeti"),
  titlePanel("California Forest Fire Analysis"),
  # set up two tab panels
  tabsetPanel(
    tabPanel("Scatter Plot", fluid = TRUE,
             sidebarLayout(
               # sidebar inputs for scatterplot
               sidebarPanel(
                 selectInput("var", "X-axis:",
                             c("Acres Burned" = "AcresBurned",
                               "# of Fatalities" = "Fatalities",
                               "# Injured" = "Injuries"),
                             selected = "AcresBurned"),
                 selectInput("var1", "Y-axis:",
                             c("Acres Burned" = "AcresBurned",
                               "# of Fatalities" = "Fatalities",
                               "# Injured" = "Injuries"),
                             selected = "Injuries"),
                 selectInput("var2", "Dot Size:",
                             c("Acres Burned" = "AcresBurned",
                               "# of Fatalities" = "Fatalities",
                               "# Injured" = "Injuries"),
                             selected = "Fatalities")
              ),
              mainPanel(
                # display scatter plot in main panel
                plotOutput("scatterplot"))
              )
    ),
    tabPanel("Bar Graph", fluid = TRUE,
             sidebarLayout(
               # sidebar inputs for the bar graph
               sidebarPanel(
                 selectInput("var3", "X-axis:",
                             c("Month" = "Month",
                               "Year" = "Year"),
                             selected = "Year"),
                 selectInput("var4", "Y-axis:",
                             c("Acres Burned" = "AcresBurned",
                               "# of Fatalities" = "Fatalities",
                               "# Injured" = "Injuries",
                               "# of Structures Damaged" = "StructuresDamaged",
                               "# of Structures Destroyed" = "StructuresDestroyed"),
                             selected = "AcresBurned")),
               mainPanel(
                 plotOutput("bargraph"))
            )
    )
  )
)
```

Inside the `server` function, Alejandro also included some extra
wrangling to the data to fit his needs. After this, he proceed to create
the scatter plot and bar graph visualizations displaying the inputs
available for selection.

```{r}
# server logic containing data prep and plotting for the app
server <- function(input, output) {
  # run Ben's data cleaning script plus further cleaning for my needs
  fire_df <- cleanFireData() %>%
    select(AcresBurned, CalFireIncident, Started, Extinguished, 
           Fatalities, Injuries, StructuresDamaged, StructuresDestroyed) %>%
    mutate(AcresBurned = as.numeric(AcresBurned),
           Started = as.Date(Started, "%Y-%m-%d"),
           Extinguished = as.Date(Extinguished, "%Y-%m-%d")) %>%
    mutate(Month = month(Started, label=TRUE), Year = year(Started)) %>%
    filter(Year != 1969) %>% # remove Year outliers found in data set
    select(-c(Started, Extinguished))
  
  # convert 0 and NAs to the mean of AcresBurned  
  fire_df$AcresBurned[fire_df$AcresBurned == 0] <- NA
  fire_df$AcresBurned[is.na(fire_df$AcresBurned)] <- mean(fire_df$AcresBurned, 
                                                          na.rm = TRUE)
  # create season variable for scatter plot color use joined with original data frame
  fire_df$Month <- as.numeric(fire_df$Month)
  
  # setting month range by seasons
  spring <- c(03,04,05)
  fall <- c(09,10,11)
  summer <- c(06,07,08)
  winter <- c(12,01,02)
  # seasons df to join with existing df
  seasons <- data.frame(
    Season = c("spring","spring","spring", "fall","fall","fall", 
               "summer","summer","summer", "winter","winter","winter"),
    Month = c(spring, fall, summer, winter)
  )
  # joining seasons by month
  fire_df <- fire_df %>%
    full_join(seasons, fire_df, by = "Month") %>%
    # setting variables to desired format
    mutate(Month = month(Month, label=TRUE),
           Season = as.factor(Season))
  
  # scatter plot showing the desired inputs
  output$scatterplot <- renderPlot(
    ggplot(fire_df, aes(x=fire_df[,input$var], y=fire_df[,input$var1])) +
      geom_point(aes(color=Season, size=fire_df[,input$var2])) +
      labs(title=paste(input$var1, "vs.", input$var),
           x=paste(input$var), y=paste(input$var1), 
           size=paste(input$var2)) +
      scale_x_continuous(labels = scales::comma)
  )
  
  # bar graph of the desired inputs
  output$bargraph <- renderPlot(
    ggplot(fire_df, aes(x=as.factor(fire_df[,input$var3]), 
                        y=fire_df[,input$var4], fill=Season)) +
      geom_bar(position="dodge", stat="identity", width=0.7) +
      labs(title=paste(input$var4, "by", input$var3),
           x=paste(input$var3), y=paste(input$var4)) +
      scale_y_continuous(labels = scales::comma)
  )
}

# set plot theme to minimal
theme_set(theme_minimal())
# have the shiny theme match the ggplot theme
thematic_shiny()
```
To view the app, click this [link](https://acepeda.shinyapps.io/WildfiresApp/).
```{r include=FALSE}
# deploy the app
#deployApp()
# Run the app
shinyApp(ui = ui, server = server)
```

The shiny app shows that there may not be a correlation with the season
how intense a California wildfire can get. Although there is a
noticeable trend during the summer season throughout the years
2013-2019, the wildfires during the fall season in 2017 says otherwise.

------------------------------------------------------------------------

#### Training a Model for Correlation:

```{r message=FALSE}
#Using my data:
library(dplyr)
library(e1071)
library(kernlab)
library(ggplot2)
library(Metrics)
#Cleaning the data to only include the columns I need
data <- read.csv("California_Fire_incidents.csv")
data <- data %>%
        select(AcresBurned, ArchiveYear, CalFireIncident, Started, MajorIncident)
data$CalFireIncident <- as.factor(data$CalFireIncident)


#Removing the the year and day from the date so that I can only use month
data$month <- strftime(data$Started,"%m")
data$day <- strftime(data$Started, "%j")
data$month <- as.numeric(data$month)
data$day <- as.numeric(data$day)

#season column
data$seasons <- as.numeric(data$month)

#Creating a separate df for seasons
spring <- c(03, 04, 05)
fall <- c(09, 10, 11)
summer <- c(06, 07, 08)
winter <- c(12, 01, 02)

seasons <- data.frame(
  seasons = c("spring","spring","spring", "fall","fall","fall", "summer","summer","summer", "winter","winter","winter"),
  month = c(spring, fall, summer, winter)
)

data <- data %>%
        full_join(seasons, data, by = "month")

```

Upon looking at the original data frame, wanting to see if there were a
correlation between the seasons and the amount of acres burned. In
theory dry conditions in the summer would cause more fires due to the
intense heat that California gets during the summer.

Using Ben's initial cleaning of the original data set, more cleaning was
needed in order for Brandon to obtain the seasons for the months:

```{r}
#Creating final dataset
data_fin <- data %>%
            select(AcresBurned, ArchiveYear, CalFireIncident, "season" = seasons.y, month, day, Started, MajorIncident) %>%
            filter(CalFireIncident == "True")
data_fin$Started <- as.Date(data_fin$Started)
data_fin$Day <- strftime(data_fin$Started,"%m-%d")
data_fin$MajorIncident <- as.factor(data_fin$MajorIncident)
data_fin$AcresBurned <- as.numeric(data_fin$AcresBurned)
data_fin$season <- as.factor(data_fin$season)
data_fin$month <- as.factor(data_fin$month)
data_fin$AcresBurned[data_fin$AcresBurned == 0] <- NA
data_fin$AcresBurned[is.na(data_fin$AcresBurned)] <- mean(data$AcresBurned, na.rm = TRUE)
head(data_fin)
```

By adding the seasons, Brandon's correlational analysis would be easier,
due to his predictive model only predicting the four seasons, rather
than the 12 months. In theory, Brandon's predictive model using a
support vector machine, if his predictive model could accurately predict
the seasons using the Acres Burned, whether the fire was considered
major, and if it was reported by the California wildfire team, an
accurate predictive model would mean that the size of the fire has a
correlation between the time of year, and in our case which season of
the year.

Using ggplot, we could find if there are any correlations to be made:

```{r}
ggplot(data_fin, aes(x = Day, 
                     y = AcresBurned, 
                     color = season)) +
                geom_point() +
                geom_jitter() +
                theme(axis.text.x=element_blank()) + 
                ylim(0, 5000) +
                labs(x = "Entire Year in Days")
```

Considering that the days are chronological in order due to the year,
the plot will be shown accordingly. With the division of the plot with
the different colors showing the different seasons, there is not
potential for a linear model to show a prediction, however a predictive
model using classification would have a better fit considering that the
column we are predicting is a factor with 4 levels: Fall, Winter,
Spring, and Summer.

Using a random sample that we learned in module 10, we can create a test
and training set which we will use for our analysis and model:

```{r echo=FALSE}
#Train and Test set
set.seed(123)
indices <- sample(nrow(data_fin), .70 * nrow(data_fin))
train <- data_fin[indices, ]
train <- train %>%
         select(AcresBurned, season, MajorIncident)

test <- data_fin[-indices,]
test <- test %>%
        select(AcresBurned, season, MajorIncident)
head(test)
head(train)
```

Going back to Brandon's analysis, the use of a support vector machine
would be used to see if there is a significance of the stated vectors in
order for the algorithm to accurately predict the season, if the model
could predict the seasons with high accuracy then there would be a
significance in season to the size of the fire.

Using the svm() function from the e1071 library, we could use the smv
algorithm to predict the seasons using classification.

```{r}
#I want to see if my model could predict the season using the variables
svm_model <- svm(season ~., data = train)
svm_model

```

```{r echo=FALSE}
#Predicting dataset to compare with test dataset
svm_pred_season <- predict(svm_model, test, cost = 100, gamma = .01)
svm_pred <- test
svm_pred$predseason <- svm_pred_season

#Checking accuracy
percentage <- sum(test$season == svm_pred$predseason) / nrow(svm_pred) * 100
sprintf("Model's Accuracy: %0.1f%%", percentage)
```

Considering that the model's accuracy is \~64%, and the predicted output
only gives out summer, the low accuracy shows that the only correlation
the model can will outfit the summer factor, meaning that there is no
correlation between the seasons and size of fire.

------------------------------------------------------------------------

#### Brandon Dora's Part:

```{r}
### Brandon Dora's Density plot of the Fires per County in California. 


### Importing Library 
library(readr)
library(ggmap)
library(ggplot2)
library(tibble)
library(tidyverse)
library(zipcode)
library(remotes)


```
### Density Plot and State Map showing fire occurences in the State of California.

For the portion below, Brandon Dora had decided to create a density map similar 
to the ones that we had created prior in week 4 of Introduction to Data Mining.

```

#implementing Ben's cleaned dataset using alejandro's inner portion of his function
 testFrame3 <- read.csv("California_Fire_Incidents.csv")
  
  testFrame3 <- testFrame[,-2]
  
  testFrame3 <- testFrame[,-6]
  
  testFrame3 <- testFrame[,-9]
  
  testFrame3 <- testFrame[,-14:-16]
  
  testFrame3 <- testFrame[,-21]
 
  testFrame3 <- testFrame[,-22]

  testFrame3 <- testFrame[,-25]
  
  summary(testFrame3$StructuresEvacuated)

  testFrame3 <- testFrame3[,-27]
  
  summary(testFrame3$StructuresThreatened)
  
  testFrame3 <- testFrame3[,-28:-29]
  
  testFrame3 <- testFrame3 %>% relocate(Started, .before = Extinguished)
  
  testFrame3 <- testFrame3 %>% relocate(Name, .before = AcresBurned)
  
  testFrame3 <- testFrame3 %>% relocate(Longitude, .before = Location)
  
  testFrame3 <- replace_na(testFrame3, list(AirTankers = 0, CrewsInvolved = 0, Dozers = 0, Engines = 0, Fatalities = 0, Helicopters = 0, Injuries = 0, PersonnelInvolved = 0, StructuresDamaged = 0, StructuresDestroyed = 0, StructuresThreatened = 0, WaterTenders = 0))


## creating and cleaning data frames to map out the state of california
  
states <- map_data("state")
California <- subset(states, region == "california")
California$Longitude <- California$long
California$Latitude <- California$lat
California$long = NULL
California$lat = NULL
California$region <- str_to_title(California$region)
counties <- map_data("county")
californiaCounties <- subset(counties, region == "california")
californiaCounties$Counties <- str_to_title(californiaCounties$subregion)
californiaCounties$subregion = NULL
californiaCounties$Longitude <- californiaCounties$long
californiaCounties$Latitude <- californiaCounties$lat
californiaCounties$long = NULL
californiaCounties$lat = NULL

### creating a new dataframe based on our original to focus on the counties, and the number of times a fire has occured in each of them. 
```
The reasoning behind creating a new dataframe for this, was to be able to seamlessly consolidate the name's of each county in California,
while also being able to log the number of times a fire had occured in each county. This way, we are able to join our dataframe, with the newly created "californiaCounties" dataframe listed above. There are many other ways to do this, but for the method he had chosen it was simply because it made for
a much more seamless process, and turned out to be more time effective. 
```


testFrame2 <- table(testFrame3$Counties)
testFrame2

testFrame2 <- as.matrix(testFrame2)
testFrame2 <- as.data.frame(testFrame2)
testFrame2$Counties <- c("Alameda", "Alpine", "Amador", "Butte", "Calaveras", "Colusa", "Contra Costa", "Del Norte", "El Dorado", "Fresno", "Glenn", "Humboldt", "Inyo", "Kern", "Kings", "Lake", "Lassen", "Los Angeles", "Madera", "Marin", "Mariposa", "Mendocino", "Merced", "Mexico", "Modoc", "Mono", "Monterey", "Napa", "Nevada", "Orange", "Placer", "Plumas", "Riverside", "Sacramento", "San Benito", "San Bernardino", "San Diego", "San Joaquin", "San Luis Obispo", "San Mateo", "Santa Barbara", "Santa Clara", "Santa Cruz", "Shasta", "Sierra", "Siskiyou", "Solano", "Sonoma", "Stanislaus", "State of Nevada", "State of Oregon", "Sutter", "Tehama", "Trinity", "Tulare", "Tuolumne", "Ventura", "Yolo", "Yuba")
testFrame2$FireCount <- testFrame2$V1
testFrame2$V1 = NULL


### combining our new test frame with our county dataframe. 
californiaCounties <- left_join(californiaCounties, testFrame2, by="Counties")
californiaCounties[is.na(californiaCounties)] = 0

```
As shown above, joining the two dataframes together allowed for us to easily combine the data we needed, while not having any complications whatsoever. 

```


# creating our state map
caliFire <- ggplot(data = California, mapping = aes(x = Longitude, y = Latitude, group = group))
caliFire <- caliFire + geom_polygon(data = californiaCounties, color = "white") + geom_polygon(color = "black", fill = NA)
caliFire <- caliFire + coord_fixed(1.3)

### the state map sectioned by counties
caliFire

```
The reasoning behind plotting a map displaying counties first, was to get a good idea of what we are working with,
while also creating a base that we will be placing our density map on top of. 
```

### adding our data and creating our 2d density map on top our previously created map.


caliFire <- caliFire + geom_polygon(data = californiaCounties, aes(fill = FireCount))
caliFire<- caliFire + ggtitle("Fire Count in California by County")
caliFire2d <- caliFire + stat_density2d()
caliFire2d 


```
As you can see, our density map is properly displaying our Two-Dimensional density map, where we can 
easily see by county where more often than not fires had occured. 
```



```

